---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
<!-- badges: end -->

# PLEASE NOTE:This package as well as this readme are still under development. 

# Welcome to disperseR

## What is disperseR?

`disperseR` is an R package designed based on the [hyspdisp package](https://github.com/lhenneman/hyspdisp) and the [SplitR package](https://rdrr.io/github/rich-iannone/SplitR/). It is very important to note that many functions in `disperseR` are just sightly redesigned functions from the two mentioned packages. 


It runs the [HYSPLIT](https://www.ready.noaa.gov/HYSPLIT.php) many times and calculates the HYSPLIT Average Dispersion (or HyADS) exposure metric
The results can then be aggregated to ZIP code level to create national estimates of exposure from various sources. 

Thanks to the [hyspdisp package](https://github.com/lhenneman/hyspdisp), for example, plumes from several power plants can be tracked for many days and cumulative impacts estimated. disperseR laverages [hyspdisp package](https://github.com/lhenneman/hyspdisp) and allows the user have a more friendly interaction with the package.

### What is improved? 

`disperseR` is a new version of the hispdisp package. What has been improved?

- Input data manipulation is handled at the package level. The user only has to read the data in using the `disperseR::get_data()` function. We show how to do it in the main vignette. 

- We also created additional vignettes should the user want to see how the attached data was preprocessed. We show every single step of preprocessing starting from the step of data download. This is key for reproducible research. 

- Very clear project struture and automatization does not make the user lost in the maze of multiple folders. The `disperseR::create_dirs()` automatically creates the whole project structure either in the specified location or on the desktop. The function also assigns path to each folder to the R environment. These paths are then used by other `disperseR` functions. **Note** that the `disperseR::create_dirs()` function does not overwrite the project folders if they already exists in the specified location. 

- Until now the `units` data for different years was separated and only four years of data were available with the package. Now data for years 1995 to 2015 has been added and aggregated to one data file called `units` attached to `disperseR`.

- ZIP code linkage procedure requires a ZCTA-to-ZIP code crosswalk file. These crosswalk data has also been attached to the package. It not only provides the crosswalk between ZCTA and ZIP but also contains information about population sizes.

- Before the user could only run analysis for one year. `disperseR` allows to process all the needed years together. 

- Graph functions now have many automatic features. 


### Vignettes attached with the package 

- [The main vignette for disperseR](vignettes/Vignette_Disperse.html)
- [Crosswalk data preparation](vignettes/Vignette_Crosswalk_Preparation/html)
- [Planetary layers data preparation](vignettes/Vignette_Planetary_Layers_Data_Preparation.html)
- [ZIP Code coordinate data preparation](vignettes/Vignette_Zip_Code_Coordinate_Data_Preparation.html)
- [Vignette_ZCTA_Shapefile_Preparation](vignettes/Vignette_ZCTA_Shapefile_Preparation.html)
- [Vignette_Units_Preparation](vignettes/Vignette_Units_Preparation.html)

### Data attached with the package 
- **crosswalk**: ZIP code linkage procedure requires a ZCTA-to-ZIP code crosswalk file. ZCTAs are not exact geographic matches to ZIP codes, and multiple groups compile and maintain Crosswalk files. We used the Crosswalk maintained by [UDS Mapper]("https://www.udsmapper.org/zcta-crosswalk.cfm") and prepossessed it also including information about the population size. While not necessary for the HYSPLIT model or processing of its outputs, population-weighted exposure metrics allow for direct comparisons between power plants. If you would like to know more details about how this crosswalk was prepared, we have attached a vignette that explains it. You can see it by clicking [here](vignettes/Vignette_Crosswalk_Preparation/html)

- PP.units.monthly1995_2017.RData (we currently do not have a vignette for these data due to server problems of the data owner)
- units1995 - units2015 
- zipcode coordinate data

## Instructions

### Download and installation. 

Start by typing the following in your R console. This will download the package from GitHub, install it and build the vignettes. 

```{r, eval = FALSE}
devtools::install_github("garbulinskamaja/disperseR", force = TRUE, build_vignettes = TRUE)
library(disperseR)
```

### Set up the project. 

The vignettes will instruct you to do so but you can already start by creating the project folder. Use `disperseR::create_dirs()` function to do so. Point `disperseR` to the location where you want your project to be created. For example the following code will create the project in the user's Dropbox. If you do not specify the location and just type `disperseR::create_dirs()` it will still work and the project will be created on your desktop. 


```{r, eval = FALSE}
disperseR::create_dirs(location="/Users/username/Dropbox")
```

This will set up is the following folders and paths to them :

* `main`: the main folder where the project will be located. 
  + `input`: the input that we need for calculations. 
    * `zcta_500k`: ZCTA (A Zip Code Tabulation Area) shape files
    * `hpbl`: monthly global planetary boundary layer files.
    * `meteo`: (reanalysis) meteorology files
  + `output`
    * `hysplit`: disperseR output (one file for each emissions event)
    * `ziplink`: files containing ZIP code linkages
    * `rdata`: RData files containing HyADS source-receptor matrices
    * `exp`: exposure per zipcode data 
    * `graph`: graphs saved here as pdf when running functions
  + `process`: temporary files that are created when the model is running and then deleted



## References / Resources Used

- NCEP Reanalysis data provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their Web site at https://www.esrl.noaa.gov/psd/

- 

Citation: Mesinger, F., G. DiMego, E. Kalnay, K. Mitchell, P.C. Shafran, W. Ebisuzaki, D. Jović, J. Woollen, E. Rogers, E.H. Berbery, M.B. Ek, Y. Fan, R. Grumbine, W. Higgins, H. Li, Y. Lin, G. Manikin, D. Parrish, and W. Shi, 2006: North American Regional Reanalysis. Bull. Amer. Meteor. Soc., 87, 343–360, https://doi.org/10.1175/BAMS-87-3-343
) 
