---
title: "Vignette Planetary Layers Data Preparation (Optional)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Planetary Layers Data Preparation (Optional)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, eval = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this vignette, we show how we prepared the monthly mean planetary boundary layer heights data that can be downloaded using the `get_data()` function specifying the `data = "pblheight"`. 

First, we load the packages that we will need for completing this task. 

```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(disperseR)
library(data.table)
library(tidyverse)
```

In case you have not yet created your project folder use the following command. 

```{r, eval = FALSE}
disperseR::create_dirs()
```

First we specify the file names and the url where we download the data from. We used the data available on the [ESRL website](https://www.esrl.noaa.gov/psd/repository/entry/get/hpbl.mon.mean.nc?entryid=synth%3Ae570c8f9-ec09-4e89-93b4-babd5651e7a9%3AL05BUlIvTW9udGhsaWVzL21vbm9sZXZlbC9ocGJsLm1vbi5tZWFuLm5j).

```{r, eval = FALSE}
directory <- hpbl_dir
file <- file.path(directory, 'hpbl.mon.mean.nc')
url <-'https://www.esrl.noaa.gov/psd/repository/entry/get/hpbl.mon.mean.nc?entryid=synth%3Ae570c8f9-ec09-4e89-93b4-babd5651e7a9%3AL05BUlIvTW9udGhsaWVzL21vbm9sZXZlbC9ocGJsLm1vbi5tZWFuLm5j'
```

If the file does not exist in the folder download it as follows:


```{r, eval = FALSE}
if (!file.exists(file)){
  download.file(url = url, destfile = file)
}
```

Before reading in, it is necessary to set the system time zone to UTC so that the dates are formatted correctly in the raster files.

```{r, eval = FALSE}
Sys.setenv(TZ = 'UTC')
hpbl_rasterin <- suppressWarnings(raster::brick(x = file, varname = 'hpbl'))
```

The following is done to fix error in the dataset. For more information please see [here]("https://stackoverflow.com/questions/56806894/raster-warning-message-in-cbindmi-vals-number-of-rows-of-result-is-not/56807318#56807318")

```{r, eval = FALSE}
crs(hpbl_rasterin) <- "+proj=lcc +x_0=5632642.22547 +y_0=4612545.65137 +lat_0=50 +lon_0=-107 +lat_1=50"
```

Now `hpbl_rasterin` is prepared and ready to use.

