---
title: "Vignette - ZCTA Shapefile Preparation (Optional)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette - ZCTA Shapefile Preparation (Optional)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this vignette, we show how we prepared the ZCTA shape data that can be downloaded using the `get_data()` function specifying the `data = "zctashapefile"`. In case you want to use different files you can follow some of these preprocessing steps. 

First, we load the packages that we will need for completing this task. 

```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(disperseR)
library(data.table)
library(tidyverse)
```

In case you have not yet created your project folder use the following command. 

```{r, eval = FALSE}
disperseR::create_dirs()
```

We use use ZCTA shapefiles from the US census website [US census website](http://www2.census.gov/geo/tiger/GENZ2017/shp/cb_2017_us_zcta510_500k.zip). It is important to note that this website sometimes fails and the download should be postponed. 

```{r, eval = FALSE}
file <- file.path(zcta_dir, 'cb_2017_us_zcta510_500k.zip')
url <- 'ftp://ftp2.census.gov/geo/tiger/GENZ2017/shp/cb_2017_us_zcta510_500k.zip'

if(!file.exists(file)){
    download.file(url = url, destfile = file)
    unzip(file, exdir = zcta_dir)
}
```

You can can load these files into your R session as follows. 

```{r, eval = FALSE}
zcta <- file.path(zcta_dir, 'cb_2017_us_zcta510_500k.shp') # define path
zcta <- raster::shapefile(x = zcta) 
```

It is recommended to transform the ZCTA shapefile to a known projection to maintain consistency throughout the allocation process. Lat-lon projections are preferred, such as the [North American Albers Equal Area Conic](https://epsg.io/102008):

```{r, eval = FALSE}
p4s <- "+proj=aea +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m"
zcta <- sp::spTransform(x = zcta, CRSobj = p4s)
```

The ZCTA shape file should now be available in your environment. 

