---
title: "Vignette - Crosswalk Preparation (Optional)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette - Crosswalk Preparation (Optional)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this vignette, we show how we prepared the crosswalk dataset that is attached to the `hyspdisp` package. In case you want to use a different crosswalk you can follow some of these preprocessing steps. 

First, we load the packages that we will need for completing this task. 

```{r, message = FALSE, warning = FALSE, , eval = FALSE}
library(disperseR)
library(data.table)
library(tidyverse)
```

In case you have not yet created your project folder use the following command. 

```{r, eval = FALSE}
disperseR::create_dirs()
```

If the file does not exist, download it from [USDMAPPER](https://www.udsmapper.org/docs/zip_to_zcta_2017.xlsx).
Then, open it. 

```{r,  eval = FALSE}
filename<- file.path(main_dir, "crosswalk.xlsx")
if(!file.exists(filename)){
  download.file("http://www.udsmapper.org/docs/zip_to_zcta_2017.xlsx", filename)
}
crosswalk <- readxl::read_excel(filename, skip = 0)
```

Next step is to rename the `ZIP_CODE`, `CITY` variables.

```{r, eval = FALSE}
crosswalk <- crosswalk %>% 
  dplyr::rename(.,ZIP=ZIP_CODE,
    CITY = PO_NAME)
```

We use the 2017 census data packaged by [American Fact Finder](https://factfinder.census.gov/bkmk/table/1.0/en/ACS/17_5YR/B01003/0100000US.86000).

We put the downloaded csv file in our main directory that has been specified in the `createdirs()` function and read it. We are also cleaning a the column names a little bit. 

```{r, eval = FALSE}
file <- file.path(main_dir, "ACS_17_5YR_B01003_with_ann.csv")
ZCTApop2017 <- read.csv(file, skip = 1, header = T)
names(ZCTApop2017) <- c('ID', 'ID2', 'GEOGRAPHY', 'TOTALESTIMATE', 'MARGINOFERROR')
```

Now, we create a ZCTA code variable by extracting the 5 last digits from the `id` variable. 

```{r, eval = FALSE}
ZCTApop2017 <- ZCTApop2017 %>% 
  dplyr::mutate(ZCTA=stringr::str_sub(ID,-5,-1))
```

Merge the data to the crosswalk. 

```{r, eval = FALSE}
crosswalk <- dplyr::inner_join(crosswalk, ZCTApop2017)
```

Keep only some of the variables.

```{r, eval = FALSE}
crosswalk<- crosswalk %>% 
  dplyr::select(ZIP, CITY, ZCTA, STATE, TOTALESTIMATE, MARGINOFERROR)
```

Transform to data.table.

```{r, eval = FALSE}
crosswalk <- data.table::data.table(crosswalk)
```

Now the crosswalk is ready to use. 


