---
title: "Vignette - EPA AMPD Data Preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this vignette, we show how we prepared the monthly power plant emissions data that we have attached to the package. 

First, we load the packages that we will need for completing this task. 

```{r, message = FALSE, warning = FALSE}
library(disperseR)
library(data.table)
library(tidyverse)
```

In case you have not yet created your project folder use the following command. 

```{r}
disperseR::create_dirs()
```

Unfortunately, these data cannot be downloaded with a script, but we have prepared bookmarks for you.  Please download the following data. Here we download it and place it in the `main_dir` folder. 

[Data 2010 - 2018](https://ampd.epa.gov/ampd/#?bookmark=12264)


[Data 2000 - 2009](https://ampd.epa.gov/ampd/#?bookmark=12265)


[Data 1980 - 1999](https://ampd.epa.gov/ampd/#?bookmark=12266)




