---
title: "Vignette - DipserseR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette - DipserseR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette you will see how to use the `disperseR` package. We will walk you through an example step by step. In this vignette we are using functions that will automatically download the needed data for you to be able to run an analysis for the United States. We also provide other vignettes that show you how we processed the data (optional).  

## The step up 

We start by importing the packages that we will use in the analysis. Make sure you have them installed. The functions in `disperseR` are based on the `data.table` package for faster computation, but this vignette contains many functions from the `tidyverse` family to illustrate another syntax. 

```{r, message = FALSE, warning = FALSE}
library(disperseR) # our package
library(ncdf4)
library(data.table)
library(tidyverse)
library(parallel)
library(sf)
library(viridis)
library(ggplot2)
library(scales)
library(ggsn)
library(gridExtra)
library(ggmap)
library(ggrepel)
```

Next, we will use the `create_dirs()` function that creates a project folder with all the necessary sub folders. We recommend that you begin by setting up your project with this function. Provide the path to where you would like your project to live in the `location` argument. By default, if you do not specify the path, this function will create a main directory on your desktop. It will also assign paths to string variables in your environment. 

This will set up is the following folders and paths to them :

* `main`: The main folder where the project will be located. 
  + `input`: the input that we need for calculations. 
    * `zcta_500k`: ZCTA (A Zip Code Tabulation Area) shape files
    * `hpbl`: Monthly global planetary boundary layer files.
    * `meteo`: (reanalysis) meteorology files
  + `output`
    * `hysplit`: disperseR output (one file for each emissions event)
    * `ziplink`: files containing ZIP code linkages
    * `rdata`: RData files containing HyADS source-receptor matrices
    * `exp`: exposure per zipcode data 
    * `graph`: graphs saved here as pdf when running functions
  + `process`: temporary files that are created when the model is running and then deleted


Note that not specifying the `location` argument will lead to a default Desktop location for the project folder. You can specify another folder by typing: `disperseR::create_dirs(location="/Users/username/projectlocation")` instead. 

```{r}
disperseR::create_dirs() 
```

## The inputs 

The next step is to get the data necessary for the analysis.  Let's start by loading data sets that are provided inside the `disperseR` package.

### The crosswalk

ZIP code linkage procedure requires a ZCTA-to-ZIP code crosswalk file. ZCTAs are not exact geographic matches to ZIP codes, and multiple groups compile and maintain Crosswalk files. We used the Crosswalk maintained by [UDS Mapper]("https://www.udsmapper.org/zcta-crosswalk.cfm") and prepossessed it also including information about the population size. While not necessary for the HYSPLIT model or processing of its outputs, population-weighted exposure metrics allow for direct comparisons between power plants. If you would like to know more details about how this crosswalk was prepared, we have attached a vignette that explains it. 

Let's load the crosswalk file and assign it to `crosswalk`

```{r}
crosswalk <- disperseR::crosswalk
```

Let's see the first rows of the dataset. 

```{r}
head(crosswalk)
```



### The monthly powerplant emissions

The `disperseR` package also includes monthly power plant emissions, load, and heat input data, accessible using the following command:

```{r}
PP.units.monthly1995_2017 <- disperseR::PP.units.monthly1995_2017
```

Let's see the first rows of the dataset. 

```{r}
head(PP.units.monthly1995_2017)
```


### ZIP code coordinates 

The `disperseR` package contains a data set with coordinates of ZIP codes. This might be useful for plotting, but it is not necessary as it will be used automatically by our plotting functions where required. If needed however, you can access it as follows: 

```{r}
zipcodecoordinate <- disperseR::zipcodecoordinate
```

Let's see the first rows of the dataset. 

```{r}
head(zipcodecoordinate)
```


### The shape files

Equally important is the ZCTA shape file. Using the `get_data()` function and setting `data = "zctashapefile"` you can download and preprocess the file from the [US census website](http://www2.census.gov/geo/tiger/GENZ2017/shp/cb_2017_us_zcta510_500k.zip) automatically. If the file already exists in the correct folder, this function will preprocess it and load it into your R environment. We have also attached a vignette explaining this preprocessing in detail. 

```{r}
zcta <- disperseR::get_data(data = "zctashapefile")
```

```{r}
zcta
```


### Monthly mean planetary boundary layer heights 

Another input that we need are the monthly mean boundary layer heights. We used the data available on the [ESRL website](https://www.esrl.noaa.gov/psd/repository/entry/get/hpbl.mon.mean.nc?entryid=synth%3Ae570c8f9-ec09-4e89-93b4-babd5651e7a9%3AL05BUlIvTW9udGhsaWVzL21vbm9sZXZlbC9ocGJsLm1vbi5tZWFuLm5j). Again, using the `get_data()` function and setting  `data = "pblheight"`. If you need more information about how these data have been preprocessed just check the special vignette that is available with the package.


```{r}
pblheight <- disperseR::get_data(data = "pblheight")
```
```{r}
pblheight
```

### The units data 

This package contains annual emissions and stack height data from [EPA's Air Markets Program Data](https://ampd.epa.gov/ampd/) and the [Energy Information Agency](https://www.eia.gov/electricity/data/eia860/) for years 2003-2012. Again, if you would like to know how these data were prepared please see the special vignette that we have attached to this package. 

Now, we need to select the power plants whose impact we will analyse. In this case, we'll use the two units in 2005 with the greatest SOx emissions, and two with the greatest SOx emissions in 2006. You are free to choose your own units. You can see that unit "3136-1" is on top of the ranking in both years. 


```{r}
unitsrun2005 <- disperseR::units %>% 
  dplyr::filter(year ==2005)%>% 
  dplyr::top_n(2, SOx) 

unitsrun2006 <- disperseR::units %>% 
  dplyr::filter(year ==2006)%>% 
  dplyr::top_n(2, SOx) 

head(unitsrun2005)
head(unitsrun2006)
```


Below append the two datasets together and transform it to `data.table` format.  

```{r}
unitsrun <- data.table::data.table(rbind(unitsrun2005, unitsrun2006)) 
```
Let's look at the data. 
```{r}
unitsrun
```

### The meteorology files

Download reanalysis meteorology files. Below we show code to test for the meteorology files needed for the present run. They will be downloaded if they are not already in the `meteo_dir` folder. The reanalysis met files are about 120 MB each.

If you, for example, you want to download files for Jan-Jul 2005, you just have to use the `get_data()` function and set `data = "metfiles"`, `start.year = "2005"`, `start.month = "01"`, `end.year = "2005"`, and `end.month = "03"`. See below.

```{r}
disperseR::get_data(data = "metfiles", 
  start.year = "2005", 
  start.month = "07", 
  end.year="2006", 
  end.month="06")
```

Now the data should be downloaded to your `meteo_dir` directory and we can start the analysis.

## Analysis

### Define time periods to be run

HYSPLIT, as applied here, tracks air parcels emitted at certain times and locations. `disperseR` refers to these as *emissions events*. Once HYSPLIT has been run for each emissions event, the simulated parcel locations are aggregated by source, time, and location. The functions below are written to enable runs of many emissions events.

To define an object that includes all emission events in a given time period, we can use the helper function `define_inputs()`. This takes as inputs a starting and ending day, and outputs a table of values whose rows will later correspond to inputs into the main `disperseR` functions. The following command combines the units defined above with four times a day for January-July in 2005 (we show an example for this short period of time here to keep the computation manageable for a desktop computer). Four daily emissions events are defined by the `start_hours` variable, and `duration = 240` denotes that the emitted air parcels are tracked for 240 hours (10 days). 10 days is the maximum (approximately) that sulfur stays in the atmosphere before it deposits to the ground. The result of running this function is a `data.table` that you can see below. 

```{r}
input_refs <- define_inputs(units = unitsrun,
  startday = '2005-07-01',
  endday = '2006-06-30',
  start.hours =  c(0, 6, 12, 18),
  duration = 240)

head(input_refs)
```

### Run HYSPLIT
The following examples show how to run a small subset (chosen to provide workable examples on a laptop) of emissions events, link to ZIP codes, and plot the results.

The `run_fac_parallel()` function runs HYSPLIT for each emissions event. Inputs defined above, including the projected ZCTA shape file (`zcta`), the ZIP-ZCTA crosswalk (`crosswalk`), the planetary boundary layer raster (`pblheight`), are necessary in the function call. With `link2zip = T` you can link dispersion patterns from individual emissions events to ZIP codes; this, however, somewhat violates the spirit of HyADS, since the large number of emissions events is used as a check on some other uncertainties introduced by the simplifying assumptions. 

**If you get an error running HYSPLIT, it is likely because the meteorology files did not download correctly. Check your `meteo_dir` to ensure all files are present and at least 100MB in size.**

Below we subset the data to run the process on a local computer. We only choose the first day and first hour of each month for each of the units that we chose. 

```{r}
input_refs_subset <- input_refs[
  format(as.Date(input_refs$start_day,
  format = "%Y-%m-%d"), 
  format = "%d") == "01" & start_hour==0]
```

Now we need to get indexes of the correct rows. In this case will be all the indexes in this subset of data. 

```{r}
run_sample <- seq(1, nrow(input_refs_subset))
```

Let's see the indexes. Here we will be running the model using all the rows of the `input_refs_subset` data subset. 

```{r}
run_sample 
```

Now we will run the fac model in parallel. The argument `sample` should take the vector of indexes that we created above, so here it will be the `sample = run_sample` vector. 


`input.refs` should be the data table that is the result of the `define_input()` function. Remember that here we are running the model just on a subset. 


The `ztca.`, `crosswalk.`, and `pbl.height` arguments are `NULL` by default but you should provide them for the function to run properly. We use the `zcta`, `crosswalk`, and `pblheight` data respectively.


The package has the possibility to use two types of species. The default one is `species = 'so2'`, but you can also use particulate sulfate `species = 'so4p'`. 


The `parallel::detectCores()` function automatically detects the number of cores on your machine. If you would like to change on how many cores the model should be run please specify the number of cores. Here, for example we specified the number of cores to be the number of the cores on the machine. If you would like to run this model sequentially please set `mc.cores = 1`.


It is possible that running the below code with output a warning "WARNING: map background file not found ../graphics/arlmap". It is safe to ignore it. 

```{r}
hysp_raw <- run_fac_parallel(sample = run_sample,
  input.refs = input_refs_subset,
  pbl.height = pblheight,
  crosswalk.= crosswalk, 
  zcta. = zcta, 
  species = 'so2',
  link2zip = FALSE, ## FALSE BY DEFAULT
  proc_dir = proc_dir,
  overwrite = FALSE, ## FALSE BY DEFAULT
  npart = 100, 
  mc.cores = parallel::detectCores())
```


### Link results to ZIP codes

Most current implementations of HyADS, instead of linking dispersion patterns from individual emissions events, link the patterns by month. 
With the `link_all_units_zip()` function, users can link all air parcels to ZIP codes by month for specified units. Here, we define the variables `yearmons` with combinations of years and months. `link_all_units_zip()` uses another function called `disperser_link_zip()` that runs in the background and reads in all the relevant files (i.e., those that correspond to the provided units) produced by the `run_fac_parallel()` function and saves them. Then it links them to ZIP codes. `link_all_units_zip()` can be set to paralelize the runs by splitting different months on different cores, but you can make it serial just by setting the `mc.cores = 1`. The result of `dipserser_link_zip()` is a data.table of ZIP codes and relative contributions `N`, and an identical `.csv` file is saved to the folder with a path specified by the `ziplink_dir` variable. `N` is not weighted by the `unit`'s emissions. 

Below we present how to use `link_all_units_zip()` for chosen units. First choose the months that you would like to run the function for. Here we only choose Jan and Jul of 2005.

```{r}
yearmons<- c(paste0(2005, 7:12),  paste0(2006, 1:6))
```
Let's look again at our `unitsrun` data set. 

```{r}
unitsrun
```

We only have three units in our `unitsrun` data set. "3136-1", "3136-2" and "3149-1". We will run `link_all_units_zip()`  for all of them.

The function will return only the values for units and months that could have been linked to one another. 

If you wanted to run `link_all_units_zip()
` just for "3149-1" for example, run the code that we commented out below. 

```{r}
#unitsrun<-unitsrun[ID=="3149-1"]
```

If you want to run the function for a specific start and end date you can specify the `start.date` and the `end.date` arguments. For example, `start.date="2005-01-02"` for 2 January 2005. These arguments are set to `NULL` by default and the function computes the start and the end dates using the `year.mons` provided. `pbl.height = pblheight` by default but you can change it. The same applies to `crosswalk. = crosswalk`. As mentioned before  `link_all_units_zip` enables the parallel run by default splitting different months on different cores, but you can make it serial just by setting the `mc.cores` argument to `1`. `duration.run.hours = 240` by default which equals 10 days. As mentioned above 10 days is the maximum (approximately) that sulfur stays in the atmosphere before it deposits to the ground.

```{r}
data_linked <- link_all_units_zip(units.run=unitsrun, 
  mc.cores = 4, 
  year.mons = yearmons,
  pbl.height = pblheight,
  crosswalk. = crosswalk,
  duration.run.hours = 240,
  overwrite = FALSE)

head(data_linked)
```

It could be the case that data could not have been linked for a specific month and unit. Below we show how to take a look at the combinations of units and months that were successfully linked. 

We can easily do it in the following way:

```{r}
unique(data_linked$comb)
```

### Visualization of the results. 

At this point, it makes sense to take a look at the results to see which ZIP codes are impacted by emissions from the facility whose impacts we modeled above. We can see how the impact looks by looking at numbers and plots. We will first just look at the numbers. For this purpose, we created a function `create_impact_table_single` that generates a specific data.table.

First we need to bring in the zcta data set as follows. 

```{r}
zcta_dataset <- disperseR::get_data("zcta_dataset")
```

And now we can use the `create_impact_table_single` function to see the computed impact for a single month and a single unit. 

Keep in mind, since we only modeled a subset of the potential model runs defined in `input_refs`, the impacts contained in `impact_table_single` and plotted below include only results from a single emissions event (is it actually a single event of single month?). Further, the plotted values are "HyADS raw exposure", and are comparable only to a relative exposure of 0. So, a value of 0.01 represents 10x more exposure than a value of 0.001, but only in relation to unit 3136-1.

```{r}
impact_table_single <- create_impact_table_single(data.linked=data_linked, 
  data.units = unitsrun,
  zcta.dataset = zcta_dataset,  
  map.unitID = "3136-1",
  map.month = "20058", 
  metric = 'N')

head(impact_table_single)
```

In a similar way we can just plot the results using the `plot_impact_single` function. A customized title can be set using the `plot.title` argument.

```{r, fig.height = 4, fig.width = 7}
ziplink_plot <- plot_impact_single(data.linked=data_linked, 
  map.unitID = "3136-1",
  map.month = "20058", 
  data.units = unitsrun,
  zcta.dataset = zcta_dataset,  
  metric = 'N',
  graph.dir = graph_dir, 
  legend.title = 'HyADS raw exposure')
  
ziplink_plot
```

### Combine all results into RData file.
So far, we have run two worker functions important to the implementation of HyADS:

`run_fac_parallel()` ran HYSPLIT for unit-date-time emission event combinations and saved an output file for each gathered output files from `run_fac_parallel()`. Then the `link_all_units_zip()` function grouped them by month and unit, and linked the HYSPLIT parcel locations contained in the files to ZIP codes.
In practice, these two worker functions may need to be implemented in parallel R sessions on a cluster to improve efficiency. It helps for future analyses to gather the relevant monthly files and save them as a single RData file. This is made possible by `combine_monthly_ziplinks()`, which saves an RData file of annual monthly results to the `rda_dir`.


```{r}
combined_ziplinks <- combine_monthly_ziplinks(month_YYYYMMs = yearmons)
names(combined_ziplinks)
```

## Calculate and extract useful information from the results
Up until now, this vignette has focused on producing unweighted HyADS results.Once the procedure above has been followed, it is fairly straightforward to weight the unweighted HyADS results by emissions (or other quantity) and aggregate the results by source, receptor, and/or time period. 

### Weight the results by emissions

The `create_impact_table_weighted()` function takes as input the .RData file created by `combine_monthly_ziplinks()` and monthly power plant emissions. The user can select the disperseR year (`year.D`), the emissions year (`year.E`), the weighting pollutant (although it does not necessarily have to be a pollutant), the source aggregation (you can choose `source.agg = "total"`, `source.agg "facility"`, or `source.agg = "unit"`), the time aggregation (`time.agg = "month"` or `time.agg = "year"`). 

Here, we first calculate the annual HyADS emissions weighted exposure for all emissions modeled above and plot the results. The plots require a large amount of working memory, and may require you to clear your image history (if using RStudio).

```{r}
zip_exp_ann <-disperseR::calculate_zip_exposure(
  rda_file = file.path(rdata_dir, "hyads_unwgted.RData"),
  units.mo = PP.units.monthly1995_2017,
  year.E = 2006,
  year.D = 2006,
  pollutant = 'SO2.tons',
  source.agg = 'total',
  time.agg = 'year',
  exp_dir = exp_dir)
```

We can plot the results as following using the `plot_impact_weighted()` function. Again, a customized title can be set using the `plot.title` argument. You can also zoom in to a specific part of the map by setting the `x.lim` and `y.lim` arguments. By default they are set to `c(-123, -69)` and `c(24, 50)` respectively. 

Please note that the produced graphwwill be automatically saved in the folder specified in the `graph.dir` argument. Here we set `graph.dir  = graph_dir`. `graph_dir` should have been automatically created by the `create_dirs()` function. 

```{r, fig.height = 4, fig.width = 7}
zip_exp_ann_plot <- disperseR::plot_impact_weighted(data.linked=zip_exp_ann,
  data.units = unitsrun,
  zcta.dataset = zcta_dataset,  
  time.agg = 'year',
  metric = 'hyads',
  legend.title = 'Aggregate HyADS exposure',  
  graph.dir = graph_dir)

zip_exp_ann_plot
```

Let's now see how to plot a monthly exposure. 

Results are saved to the `exp_dir` and returned, but `return.monthly.data` must be set to true to return data when `time.agg = 'month'` (the default was set to save working memory space).

```{r}
zip_exp_mon <- disperseR::calculate_zip_exposure(rda_file = file.path(rdata_dir, "hyads_unwgted.RData"),
  units.mo = PP.units.monthly1995_2017,
  year.E = 2005,
  year.D = 2005,
  pollutant = 'SO2.tons',
  source.agg = 'total',
  time.agg = 'month',
  exp_dir = exp_dir, 
  return.monthly.data = T)
```

Here we also illustrate how to use the limit arguments to zoom in on the map. By setting `x.lim = c(-66, -84)`, for example, we restrict the map only to the map to display only points within the 65$^{\circ}$W and 84$^{\circ}$W longitude points. Latitude limits are set by `y.lim = c(36, 44)`. 

```{r, fig.height = 4, fig.width = 7}
zip_exp_mon_plot <- disperseR::plot_impact_weighted(data.linked = zip_exp_mon,
  data.units = unitsrun,
  zcta.dataset = zcta_dataset,  
  time.agg = 'month',
  map.month = "20059", 
  metric = 'hyads',
  legend.title = 'Montly HyADS exposure', 
  x.lim = c(-66, -84), # specify coordinates here 
  y.lim = c(36, 44),
  graph.dir = graph_dir)

zip_exp_mon_plot
```

### Plot unit-specific impacts over time

Next, we calculate the monthly emissions weighted HyADS exposure (again, only from the select few runs we initiated above). The code below plots the monthly exposure contributions of the three units in ZIP codes 17815, 08210, 13849, 17229. The location of the zip codes can be seen on the small map on the right side of the plot. Please note that the axis values are different for different ZIP codes. 

```{r}
zip_exp_unit_mon2005 <- disperseR::calculate_zip_exposure(rda_file = file.path(rdata_dir, "hyads_unwgted.RData"),
                                     units.mo = PP.units.monthly1995_2017,
                                     year.E = 2005,
                                     year.D = 2005,
                                     pollutant = 'SO2.tons',
                                     source.agg = 'unit', # note! 
                                     time.agg = 'month',
                                     exp_dir = exp_dir,
                                     return.monthly.data = T)


zip_exp_unit_mon2006 <- disperseR::calculate_zip_exposure(rda_file = file.path(rdata_dir, "hyads_unwgted.RData"),
                                     units.mo = PP.units.monthly1995_2017,
                                     year.E = 2006,
                                     year.D = 2006,
                                     pollutant = 'SO2.tons',
                                     source.agg = 'unit', # note! 
                                     time.agg = 'month',
                                     exp_dir = exp_dir,
                                     return.monthly.data = T)

zip_exp_unit_mon <- rbind(zip_exp_unit_mon2005, zip_exp_unit_mon2006)
```

```{r, fig.height = 4, fig.width = 7}
zipcodes <- c(17815, 08210, 13849, 17229)

zip_exp_unit <- disperseR::plot_impact_unit(data.linked = zip_exp_unit_mon, 
  zip.codes = zipcodes,
  graph.dir = graph_dir)
```

### Rank facilities. 

In the examples provided above, we selected coal units a priori to analyze. However, a benefit of HyADS over more complex models is the ability to easily rank-order emissions sources to identify the sources that contribute the most to exposure in a given area. 

The following code snippet applies the `calc_zip_exposure()` function to calculate annual impacts by unit (here, only the two units selected above), then rank-orders their population-weighted exposures on Pennsylvania using the `rankfacs_by_popwgt_location()` function. 

```{r}
zip_exp_ann_unit <- disperseR::calculate_zip_exposure(year.E = 2005,
                                      year.D = 2005,
                                      pollutant = 'SO2.tons',
                                      rda_file = file.path( rdata_dir, 
                                                            "hyads_unwgted.RData"),
                                      exp_dir = exp_dir,
                                      units.mo = PP.units.monthly1995_2017,
                                      source.agg = 'unit',
                                      time.agg = 'year')
# add column year for proper merge
zip_exp_ann_unit[, year := 2005]
```

In this function, the input ZIP-code linkage dataset is defined by  `zip_exp_ann_unit`; the `rank.by` argument defines the variables in `zip_exp_ann_unit` to rank by—it is possible to define multiple values (e.g., rank.by = c( 'hyads', 'SOx')) to rank by multiple ranking variables if they are available; finally, `state.value`, `city.value`, and `zip.value` can be defined to limit the area of influence (here, we only rank these units’ influences on Pennsylvania with state.value = 'PA').


```{r}
unitRanks2005 <- rankfacs_by_popwgt_location(data.linked = zip_exp_ann_unit,
  crosswalk. = crosswalk,
  rank.by = c('hyads'),
  state.value = 'PA', 
  year = 2005)

unitRanks2005
```


### Plot ranked facilities.

`plotUnitsRanked()` works very similary to other plotting function in `disperseR`. Choose the dataset of ranked facilities as well as year. 

```{r, fig.height = 4, fig.width = 7}
plotUnitsRanked <- plot_units_ranked(data.units = unitsrun, 
  data.ranked=unitRanks2005,
  year=2005, 
  graph.dir = graph_dir)

plotUnitsRanked
```

The result of this function 


